Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-11-08 inference.tex
Edits to fit to 31 pages.

\paragraph{{\sc Proof Sketch.}} 
% We prove a slightly stronger statement :
% For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
%   \begin{smathpar}
%     \begin{array}{c}
%     \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
%     I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
%     \end{array}
%   \end{smathpar}
The proof follows by structural induction on $c$. Assume that $P
\Leftrightarrow \lambda(\stl,\stg).~\stl=\Fx(\stg) \conj I(\stg)$ and $Q
\Leftrightarrow \lambda(\stl,\stg).\stl = \Fx(\stg) \cup \F(\stg)$. For each type
of $c$, our strategy is to show the premises of the corresponding
RG-rule. The base cases correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE}
statements. In all these cases, the form of $\F$ directly corresponds
to the condition linking the pre-condition $P$ and post-condition $Q$
in the premises of the RG-rules. If $\stable(\R, \Fx[\F])$, then
$\stabilize{\Fx[\F]} = \F$ and hence the proof directly follows. Otherwise,
$\stabilize{\Fx[\F]} =$ \\%\hspace{2cm}
 $\lambda (\stg).\existsl(\stg', I(\stg'),
\F( \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
$I(\stg)$, any state which obeys $P$ provides the existential $\stg'$
in the above set expression, and hence $\F$ applied on that state
matches the condition linking $P$ and $Q$. The stability of $P$ with
respect to $\R$ follows directly from $\stable(\R, I)$ and $\stable(\R, \Fx)$.

The proofs for \C{SELECT}, sequencing, and conditionals use the
inductive hypothesis to infer the RG-judgments present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = \Fx(\stg) \cup (y \bind (\lambda
z.~\F(\stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = \Fx(\stg)$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$\F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below ($\Fempty =
\lambda(\stg).~\emptyset$ denotes an empty context):
\begin{smathpar}
\begin{array}{c}
\RULE
{
   \stable(R,\I)\spc
   \stable(R,I)\\
   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc
   \Fempty \vdash c \Longrightarrow_{\langle i,\R_e,I \rangle}\F \\
   \stable(\R_c,\inctxt{\Fempty}{\F}) \spc
   \forall \stg.~ G(\stg,\F(\stg)) \spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
% \RULE
% {
%   \stable(R,\I)\spc
%   \stable(R,I)\spc
%   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
%   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc 
%    \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
%   \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
%     G(\stg, \stl \rhd \stg)\spc
%   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
% }
% {
%   \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
% }
% \end{array}
% \end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $G$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
translates an $\SL$ expression ($s$) to FOL. The translation is based
on encoding a set of element type $T$ as a unary predicate on $T$.
The predicate is represented as a meta function that accepts an $x:T$
and returns a quantifier-free proposition that evaluates to true
($\top$) if and only if $x$ is present in the set. Alternatively,
the translation may also encode the set as a predicate in the logic
itself, in which case a quantified proposition constraining the
predicate is also generated. For instance, consider the set $\{1,2\}$.
The predicate describing the set can be encoded as the function
$\lambda \v. \v=1 \vee \v=2$, with no further constraints, or it can
be encoded as the function $\lambda \v. g(\v)$ with an associated
constraint ($\phi\in\Prop^1$) defining the uninterpreted boolean
function (i.e., uninterpreted predicate) $g$:
\begin{smathpar}
  \begin{array}{lcl}
    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
  \end{array}
\end{smathpar}
The translation adopts one or the other approach, depending on the
need. For uniformity, we consider the encoding of a set as pair
($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
formula constraining any uninterpreted predicates used in $G$.

>>>>

\paragraph{{\sc Proof Sketch.}} 
% We prove a slightly stronger statement :
% For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
%   \begin{smathpar}
%     \begin{array}{c}
%     \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
%     I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
%     \end{array}
%   \end{smathpar}
The proof follows by structural induction on $c$. Let $P
=\lambda(\stl,\stg).~\stl=\Fx(\stg) \conj I(\stg)$ and $Q
=\lambda(\stl,\stg).\stl = \Fx(\stg) \cup \F(\stg)$. The base cases
correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE} statements, where
the proof is straightforward. The proofs for \C{SELECT}, sequencing,
and conditionals use the inductive hypothesis to infer the
RG-judgments present in the premises of their corresponding RG-rules.
The interesting case is the \C{FOREACH} statement, for which we use
the loop invariant $\psi(\stl, \stg) \Leftrightarrow \stl = \Fx(\stg)
\cup (y \bind (\lambda z.~\F(\stg)))$, (where assuming that $c$ is the
body of the loop, $c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = \Fx(\stg)$
which is implied by the pre-condition $P$. From the inductive
hypothesis, we can infer that each iteration corresponds to the
application of $\F$. Since all iterations are assumed to be
independent of each other, and $z$ is bound to a record in $x$ for
each iteration, we conclude that at the end of every iteration, the
loop invariant $[y \cup \{z\}/y]\psi$ will be satisfied.

@@

Theorem~\ref{thm:inference-sound} lets us replace the local judgment
of the \rulelabel{RG-Txn} rule (Fig.~\ref{fig:rg-rules}) by a state
transformer inference judgment. The soundness of a transaction's
guarantee can now be established w.r.t the effect $\F$ of the body.
The \rulelabel{RG-Txn} rule so updated is shown below ($\Fempty =
\lambda(\stg).~\emptyset$ denotes an empty context):
\begin{smathpar}
\begin{array}{c}
\RULE
{
   \stable(R,\I)\spc
   \stable(R,I)\spc
   \R_e = R \backslash \I_e \spc 
   \R_c = R \backslash \I_c \spc
   \Fempty \vdash c \Longrightarrow_{\langle i,\R_e,I \rangle}\F \\
   \stable(\R_c,\inctxt{\Fempty}{\F}) \spc
   \forall \stg.~ G(\stg,\F(\stg)) \spc
   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
}
{
  \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
}
\end{array}
\end{smathpar}
% \begin{smathpar}
% \begin{array}{c}
% \RULE
% {
%   \stable(R,\I)\spc
%   \stable(R,I)\spc
%   P(\stl,\stg) \Leftrightarrow \stl=\emptyset \wedge I(\stg)\\
%   \R_e = R \backslash \I_e \spc \R_c = R \backslash \I_c \spc 
%    \R_e \vdash \rg{P}{c}{Q} \spc \stable(\R_c,Q) \\ 
%   \forall \stl,\stg.~ Q(\stl,\stg) \Rightarrow 
%     G(\stg, \stl \rhd \stg)\spc
%   \forall \stg,\stg'.~I(\stg) \wedge G(\stg,\stg') \Rightarrow I(\stg')\\
% }
% {
%   \rg{I,R}{\ctxn{i}{\I}{c}}{G,I}
% }
% \end{array}
% \end{smathpar}
Automating the application of the \rulelabel{RG-Txn} rule for a
transaction requires automating the multiple implication checks in
the premise. While $R$, $G$, $\I$ and $I$ are formulas in
first-order logic (FOL) with a relatively simple structure, $\F$
is an expression in the set language $\SL$
(Fig.~\ref{fig:logic-syntax}) with a possibly complex structure.
Fortunately, however, there exists a semantics-preserving translation
from $\SL$ to a restricted subset of first-order logic (FOL) that
lends itself to automatic reasoning.

The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
translates an $\SL$ expression ($s$) to FOL. The translation is based
on encoding a set of element type $T$ as a unary predicate on $T$.
The predicate is represented as a meta function that accepts an $x:T$
and returns a quantifier-free proposition that evaluates to true
($\top$) if and only if $x$ is present in the set. Alternatively,
the translation may also encode the set as a predicate in the logic
itself, in which case a quantified proposition constraining the
predicate is also generated. For instance, consider the set $\{1,2\}$.
The predicate describing the set can be encoded as the function
$\lambda \v. \v=1 \vee \v=2$, with no further constraints, or it can
be encoded as the function $\lambda \v. g(\v)$ with an associated
constraint, $\phi\in\Prop^1 = \forall \nu.~g(\nu) \Leftrightarrow
\nu=1 \vee \nu=2$, defining the uninterpreted predicate $g$.
The translation adopts one or the other approach, depending on the
need. For uniformity, we consider the encoding of a set as pair
($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
formula constraining any uninterpreted predicates used in $G$.

### 2017-11-08 motivation.tex
Edits to fit to 31 pages.

%%%SJ: Reviewers may not understand what primary and foreign keys are,
%%%or why they are important.  There are also some fields (e.g., ol_i_id)
%%%that are not described either in the caption or the text.

% \begin{figure}[!t]
% \includegraphics[scale=0.5]{Figures/schema}
% \caption{Database schema of TPC-C's order management system.
%   Columns against highlighted background are primary keys. Arrows denote
%   foreign key relationships.}
% \label{fig:schema}
% \end{figure}

\begin{figure}[t]
  \centering
	\begin{subfigure}{0.4\textwidth}
	  \includegraphics[width=0.9\textwidth]{Figures/schema1}
    \caption{A valid TPC-C database. The only existing order belongs
      to the district with \C{d\_id}=11. Its id (\C{o\_id}) is one
      less than the district's \C{d\_next\_o\_id}, and its order
      count (\C{o\_ol\_cnt}) is equal to the number of order line
      records whose \C{ol\_o\_id} is equal to the order's id.  }
		\label{fig:tpcc_db1}
	\end{subfigure}
  \hspace*{.2in}
	\begin{subfigure}{0.52\textwidth}
		\includegraphics[width=0.9\textwidth]{Figures/schema2}
    \caption{The database in Fig.~\ref{fig:tpcc_db1} after correctly
      executing a \C{new\_order} transaction. A new order record is
      added whose \C{o\_id} is equal to the \C{d\_next\_o\_id} from
      Fig.~\ref{fig:tpcc_db1}. The district's \C{d\_next\_o\_id} is
      incremented. The order's \C{o\_ol\_cnt} is 2, reflecting the
      actual number of order line records whose \C{ol\_o\_id} is equal
      to the order's id (2).}
		\label{fig:tpcc_db2}
   	\end{subfigure}
\caption{Database schema of TPC-C's order management system.
  The naming
  convention indicates primary keys and foreign keys. For e.g.,
  \C{ol\_id} is the primary key column of the order line table,
  whereas \C{ol\_o\_id} is a foreign key that refers to the \C{o\_id}
  column of the order table.}
\label{fig:schema}
\end{figure}

@@

@@

\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits before $T_1$. Nonetheless, both transactions read the same value for
  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
  with the same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundness of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}.  To illustrate, consider the
behavior of the \C{new\_order} transaction when executed under a
\emph{Read Committed} (RC) isolation level, the default isolation
level in 8 of the 18 databases studied in~\cite{bailishotos}.
An executing RC transaction is isolated from \emph{dirty writes},
i.e., writes of uncommitted transactions, but is allowed to witness
the writes of concurrent transactions as soon as they are committed.
Thus, with two concurrent instances of the \C{new\_order} transaction
(call them $T_1$ and $T_2$), both concurrently placing new orders for
different customers under the same district (\C{d\_id}), RC isolation
allows the execution shown in Fig.~\ref{fig:new_order_exec}.

>>>>

@@

\begin{figure}[!t]
  \centering
	\begin{subfigure}{0.4\textwidth}
	  \includegraphics[width=0.9\textwidth]{Figures/schema1}
    \caption{A valid TPC-C database. The only existing order belongs
      to the district with \C{d\_id}=11. Its id (\C{o\_id}) is one
      less than the district's \C{d\_next\_o\_id}, and its order
      count (\C{o\_ol\_cnt}) is equal to the number of order line
      records whose \C{ol\_o\_id} is equal to the order's id.  }
		\label{fig:tpcc_db1}
	\end{subfigure}
  \hspace*{.2in}
	\begin{subfigure}{0.52\textwidth}
		\includegraphics[width=0.9\textwidth]{Figures/schema2}
    \caption{The database in Fig.~\ref{fig:tpcc_db1} after correctly
      executing a \C{new\_order} transaction. A new order record is
      added whose \C{o\_id} is equal to the \C{d\_next\_o\_id} from
      Fig.~\ref{fig:tpcc_db1}. The district's \C{d\_next\_o\_id} is
      incremented. The order's \C{o\_ol\_cnt} is 2, reflecting the
      actual number of order line records whose \C{ol\_o\_id} is equal
      to the order's id (2).}
		\label{fig:tpcc_db2}
   	\end{subfigure}
\caption{Database schema of TPC-C's order management system.
  The naming
  convention indicates primary keys and foreign keys. For e.g.,
  \C{ol\_id} is the primary key column of the order line table,
  whereas \C{ol\_o\_id} is a foreign key that refers to the \C{o\_id}
  column of the order table.}
\label{fig:schema}
\end{figure}

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundness of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes,
essentially yielding serial behaviors.  Under weak
isolation\footnote{Weak isolation does not violate atomicity as long
  as the witnessed effects are those of committed transactions},
however, interferences of various kinds are permitted, leading to
executions superficially similar to executions permitted by concurrent
(racy) programs~\cite{GHE15,HPQ+15}. 
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits before $T_1$. Nonetheless, both transactions read the same value for
  \C{d\_next\_o\_id} resulting in them  adding \C{Order} records
  with the same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
To illustrate, consider the
behavior of the \C{new\_order} transaction when executed under a
\emph{Read Committed} (RC) isolation level, the default isolation
level in 8 of the 18 databases studied in~\cite{bailishotos}.
An executing RC transaction is isolated from \emph{dirty writes},
i.e., writes of uncommitted transactions, but is allowed to witness
the writes of concurrent transactions as soon as they are committed.
Thus, with two concurrent instances of the \C{new\_order} transaction
(call them $T_1$ and $T_2$), both concurrently placing new orders for
different customers under the same district (\C{d\_id}), RC isolation
allows the execution shown in Fig.~\ref{fig:new_order_exec}.
